\chapter{References}
\label{ch:references}

\begin{enumerate}
\item Indian Roads Congress, \textit{IRC:106-1990: Guidelines for Capacity of Urban Roads in Plain Areas}. New Delhi, India: Indian Roads Congress, Jamnagar House, Shahjahan Road, New Delhi-110011, Nov. 1990, reprinted Mar. 2016. [Online]. Available: \url{https://law.resource.org/pub/in/bis/irc/irc.gov.in.106.1990.pdf}

\item Indian Roads Congress, \textit{IRC SP:41: Guidelines for the Design of At-Grade Intersections in Rural and Urban Areas}. New Delhi, India: Indian Roads Congress. [Online]. Available: \url{https://thelibraryofcivilengineer.wordpress.com/wp-content/uploads/2015/09/irc-sp-41.pdf}

\item Indian Roads Congress, \textit{IRC:93-1985: Guidelines on Design and Installation of Road Traffic Signals}. New Delhi, India: Indian Roads Congress, 1985.

\item Indian Roads Congress, \textit{IRC:86-1983: Geometric Design Standards for Urban Roads in Plains}. New Delhi, India: Indian Roads Congress, 1983.

\item F. V. Webster, \textit{Traffic Signal Settings}, Road Research Technical Paper No. 39. London, UK: Road Research Laboratory, HMSO, 1958.

\item Ultralytics, \textit{YOLOv8: State-of-the-Art Real-Time Object Detection}. Ultralytics, 2023. [Online]. Available: \url{https://github.com/ultralytics/ultralytics}

\item P. A. Lopez et al., ``Microscopic Traffic Simulation using SUMO,'' in \textit{Proc. IEEE Int. Conf. Intelligent Transportation Systems (ITSC)}, 2018. [Online]. Available: \url{https://www.eclipse.org/sumo/}

\item L. Breiman, ``Random Forests,'' \textit{Machine Learning}, vol. 45, no. 1, pp. 5-32, 2001.

\item T. Hastie, R. Tibshirani, and J. Friedman, \textit{The Elements of Statistical Learning: Data Mining, Inference, and Prediction}, 2nd ed. New York, NY, USA: Springer, 2009.

\item L. R. Kadiyali, \textit{Traffic Engineering and Transport Planning}, 7th ed. New Delhi, India: Khanna Publishers, 2013.

\item Streamlit Inc., \textit{Streamlit: The Fastest Way to Build Data Apps}. [Online]. Available: \url{https://streamlit.io/}

\item G. Van Rossum and F. L. Drake, \textit{Python 3 Reference Manual}. Scotts Valley, CA, USA: CreateSpace, 2009.

\item G. Bradski, ``The OpenCV Library,'' \textit{Dr. Dobb's Journal of Software Tools}, 2000.

\item Plotly Technologies Inc., \textit{Plotly Python Graphing Library}. [Online]. Available: \url{https://plotly.com/python/}

\item W. McKinney, ``Data Structures for Statistical Computing in Python,'' in \textit{Proc. 9th Python in Science Conf.}, 2010, pp. 56-61.

\item C. R. Harris et al., ``Array Programming with NumPy,'' \textit{Nature}, vol. 585, pp. 357-362, 2020.

\item F. Pedregosa et al., ``Scikit-learn: Machine Learning in Python,'' \textit{Journal of Machine Learning Research}, vol. 12, pp. 2825-2830, 2011.

\item C. S. Papacostas and P. D. Prevedouros, \textit{Transportation Engineering and Planning}, 3rd ed. Upper Saddle River, NJ, USA: Prentice Hall, 2001.

\item J. M. Sussman, \textit{Perspectives on Intelligent Transportation Systems (ITS)}. New York, NY, USA: Springer, 2005.

\item R. P. Roess, E. S. Prassas, and W. R. McShane, \textit{Traffic Engineering}, 4th ed. Upper Saddle River, NJ, USA: Pearson Prentice Hall, 2011.
\end{enumerate}

\section*{Web Resources}

\subsection*{Official Standards and Guidelines}
\begin{itemize}
    \item Indian Roads Congress Official Portal: \url{http://www.irc.org.in/}
    \item IRC Standards Repository: \url{https://law.resource.org/pub/in/bis/irc/}
    \item Ministry of Road Transport \& Highways: \url{https://morth.nic.in/}
\end{itemize}

\subsection*{Software and Tools}
\begin{itemize}
    \item SUMO (Simulation of Urban Mobility): \url{https://www.eclipse.org/sumo/}
    \item YOLOv8 Documentation: \url{https://docs.ultralytics.com/}
    \item Streamlit Documentation: \url{https://docs.streamlit.io/}
    \item Python Official Documentation: \url{https://docs.python.org/3/}
\end{itemize}

\subsection*{Project Repository}
\begin{itemize}
    \item GitHub Repository: \url{https://github.com/vidyasj18/Signaloptimiser}
    \item Synthetic Training Dataset: \url{https://github.com/vidyasj18/Signaloptimiser/blob/main/data/synthetic_training_dataset.csv}
\end{itemize}

\section*{AI Tools and Software Acknowledgments}

This project utilized AI-assisted tools for development and documentation purposes:

\begin{itemize}
    \item \textbf{Cursor IDE:} Used for AI-assisted code development, debugging, help for latex generation and code optimization throughout the project implementation.
    \item \textbf{Grammarly:} Used for grammar paraphasing, correction and language refinement in the technical documentation and report writing.
\end{itemize}

These tools were employed to enhance code quality and documentation clarity while maintaining full authorial control over all technical content, methodologies, and research findings presented in this work.

\newpage